Cultural Transmission and Technological Transitions during the Late Paleolithic in Korea

### 2023-02-10 analysis/paper/paper.qmd
finish editing conclusion

We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error). This is indicated by relatively high CV values of around and above `r threshold_value` and mostly low correlation coefficients of less than `r threshold_value_cor`.

A social context that favoured trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea. This model proposes that people developed stemmed points by experimenting with existing blade technologies. Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values at 0.5 and higher in chronological phase 3, after 25 ka. This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. This may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model. We might speculate that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure <!-- GP to find something relevant by Fitzhugh to cite here -->. With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.

In comparing sites with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive forms of cultural transmission of tool-making at different sites. Correlation coefficient values across the sites affirm the overall chronological trend indicating indirect bias in phase three. Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Wolpyeng and Yonsangdong that date to this period. One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, higher mobility may have resulted in more frequent contacts and connections with members of social networks. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.

>>>>

We conclude that the dominant mode of cultural transmission for the earliest stemmed points was guided variation (trial and error). This is indicated by mostly low correlation coefficients of less than `r threshold_value_cor` and relatively high CV values of around and above `r threshold_value`. A social context that favored trial and error for artifact making is consistent with Seong's *in situ* model for the appearance of stemmed points in Korea. This model proposes that people developed stemmed points by experimenting with existing blade technologies. Our results support experimentation and trial and error as important processes in the cultural transmission of this artifact technology.

The correlation coefficients suggest a change in the dominant bias of cultural transmission over time, with values around `r threshold_value_cor` in chronological phase 3, after 25 ka. This is in the same range of correlation coefficients that Bettinger and Eerkens interpreted as indirect bias. We also confirm a slight decrease in CV values in four attributes. This results may suggest a shift in the social context of transmission of tool-making skills away from trial and error and towards increased reliance on copying a model. We might speculate based on the evolutionary ecological assumption that cooler conditions of the Last Glacial Maximum promoted greater integration of social networks to buffer risks of resource failure [@fitzhugh2001risk]. With more frequent contacts and connections with members of the social network, there may have been more opportunities to learn point-making directly from socially successful individuals.

In comparing assemblages with more than five stemmed points, we found complex variation in CV values that may suggest mixtures of biases in the transmission of tool-making techniques. We did not find strong support for distinctive forms of cultural transmission of tool-making at different assemblages. Correlation coefficient values across the assemblages affirm the overall chronological trend indicating indirect bias in phase three. Raw materials appear to be important in driving this trend, with acid volcanic rocks and porphyry becoming more abundant in phase 3, and dominating assemblages at Yonsangdong that date to this period. One possibility is that these raw materials were part of an adaptive shift during the Last Glacial Maximum, perhaps the cooler climate favored an expanded range for searching for raw materials, leading to a higher diversity. Similarly, change in occupation and mobility patterns may have resulted in more frequent contacts and connections with members of social networks [@park2022did]. This may have provided more opportunities to learn point-making directly from socially successful individuals, and increased the contribution of indirect bias during cultural transmission during phase 3.
